Electricity Markets

### 2022-10-05 presentations/power_slides_law.tex
fixing charts and deck

\documentclass{beamer}

@@

@@

\begin{frame}{Why aspiring lawyers should care about electricity and what you need to know}
\begin{itemize}
\setlength\itemsep{.5em}
\item Regulatory complexity:
\begin{itemize}
\setlength\itemsep{.5em}
    \item Complex regulation, even in \textit{deregulated} jurisdictions
    \item Each part of the market has its own set of regulatory constructs (distribution, transmission, generation, retail)
\end{itemize}
\item New technology means evolving market structures:
\begin{itemize}
\setlength\itemsep{.5em}
\item Electricity is, arguably, changing faster than any other energy market
\item Alberta's electricity market in the midst of a period of market- and regulatory-driven transition
\end{itemize}
\item Economics 101 in action
\begin{itemize}
\setlength\itemsep{.5em}
\item Supply and demand curves determine prices every hour in Alberta's power market
\end{itemize}
\item The next constitutional battleground?
\begin{itemize}
\setlength\itemsep{.5em}
\item Fed/prov issues, environmental policy, section 125 jurisprudence, the oddity that is section 92(10) including the declaratory power and POGG all come into play!
\end{itemize}
\end{itemize}
\vfill
\end{frame}

@@

\begin{tabular}{p{\linewidth}}
    \centering
    \includegraphics[width=.9\linewidth]{../images/grid_storage.png} \\[\abovecaptionskip]
  Source: \url{http://www.mpoweruk.com/grid_storage.htm}
\end{tabular}

@@

\begin{frame}{Market Participants}
\begin{itemize}
\setlength\itemsep{1.7em}
\item Generation

@@

\begin{frame}{Market Regulation in Alberta}
\begin{itemize}
\setlength\itemsep{1.7em}
\item Generation is a competitive market competitive

@@

\item Distribution (wires) is regulated on a cost-of-service basis, but retail (billing) is competitive.

@@

\begin{frame}{Energy units - electricity}
\begin{itemize}
\setlength\itemsep{.75em}
\item Watts: measure of capacity (instantaneous production, installed capacity, or instantaneous demand)
\begin{itemize}
\setlength\itemsep{.5em}
\item Alberta system demand: 7,200-10,700 MW (million watts)
\item Capital Power's Genessee 3 power plant has a nameplate capacity of 450 MW
\end{itemize}
\item Watt hours: measure of energy (production or demand during a given period of time; i.e. flow through)
\begin{itemize}
\setlength\itemsep{.5em}
\item Production over a day, week, month, year
\end{itemize}
\item Volts: measure of the electrical potential or the ability to convert charge to power (Watts=amps x volts)
\begin{itemize}
\setlength\itemsep{.5em}
\item Transmission lines: 150-765 kV
\item Distribution lines: 13,800 Volts
\item Household wiring: 120-240 Volts
\end{itemize}
\end{itemize}
\vfill \end{frame}

@@

\begin{frame}{Regulatory characteristics}
\begin{itemize}
\setlength\itemsep{.25em}
\item Rate-regulated or or state-owned utilities
\item Competitive markets
\begin{itemize}
\setlength\itemsep{.15em}
\item Energy only markets: ERCOT and Alberta
\item Energy and capacity markets: MISO, PJM
\item Real-time vs day-ahead prices: PJM and others have day-ahead market and then a real-time differences market
\item Many other design characteristic differences between restructured or competitive markets
\end{itemize}
\end{itemize}
\vfill
\end{frame}

\begin{frame}{Alberta Market Design}
\begin{itemize}
\setlength\itemsep{.25em}
\item Energy-only market
\item Real time, spot pricing, no day-ahead market
\item Single node
\item Capacity market to be added in the near future
\item Transmission
\item Congestion free (no nodal pricing)
\item No transmission rights
\item Ancillary services: separate, competitive market for operating reserves, transmission-must-run, load-shed and black start
\end{itemize}

@@

\begin{tabular}{p{\linewidth}}
    \centering
    \includegraphics[width=\linewidth]{../images/cali_nodes.png} \\[.1\abovecaptionskip]
  Source: CAISO
\end{tabular}
\vfill
\end{frame}

@@

\begin{frame}{The Merit Order}
   \tikz [remember picture,overlay]
    \node[yshift=-.5cm,xshift=0cm] at (current page.center)
        {\includegraphics[width=.9\paperwidth]{../images/merit_type.png}}; \vspace{1cm}
   \vfill
\end{frame}

\begin{frame}{The Merit Order}
   \tikz [remember picture,overlay]
    \node[yshift=-.5cm,xshift=0cm] at (current page.center)
        {\includegraphics[width=.9\paperwidth]{../images/merit_offer.png}}; \vspace{1cm}
   \vfill
\end{frame}

\begin{frame}{Offers}
   \tikz [remember picture,overlay]
    \node[yshift=-.5cm,xshift=0cm] at (current page.center)
        {\includegraphics[width=.9\paperwidth]{../images/bids.png}}; \vspace{1cm}
   \vfill
\end{frame}

\begin{frame}{Hourly Loads}
   \tikz [remember picture,overlay]
    \node[yshift=-.5cm,xshift=0cm] at (current page.center)
        {\includegraphics[width=.9\paperwidth]{../images/hourly-loads.png}}; \vspace{1cm}
   \vfill
\end{frame}

\begin{frame}{Peak and Average Loads}
   \tikz [remember picture,overlay]
    \node[yshift=-.5cm,xshift=0cm] at (current page.center)
        {\includegraphics[width=.9\paperwidth]{../images/loads_clean.png}}; \vspace{1cm}
   \vfill
\end{frame}

\begin{frame}{Hourly Prices}
   \tikz [remember picture,overlay]
    \node[yshift=-.5cm,xshift=0cm] at (current page.center)
        {\includegraphics[width=.9\paperwidth]{../images/hourly_prices_col.png}}; \vspace{1cm}
   \vfill
\end{frame}

\begin{frame}{Prices over time}
   \tikz [remember picture,overlay]
    \node[yshift=-.5cm,xshift=0cm] at (current page.center)
        {\includegraphics[width=.9\paperwidth]{../images/peak_prices.png}}; \vspace{1cm}
   \vfill
\end{frame}

\begin{frame}{Forward Markets}
   \tikz [remember picture,overlay]
    \node[yshift=-.5cm,xshift=0cm] at (current page.center)
        {\includegraphics[width=.9\paperwidth]{../images/forwards.png}}; \vspace{1cm}
   \vfill
\end{frame}

@@

\begin{frame}{The Balancing Pool: What on earth does/did it do?}
\vspace{-.2cm}\centering
\hspace{-1.5cm}\begin{tabular}{p{1.1\linewidth}}
    \centering
    \includegraphics[width=.9\linewidth]{../images/balancingpool.png} \\[\abovecaptionskip]
  Source: Capital Power
\end{tabular}

@@

\begin{frame}{Costs of New Capacity Additions}
   \tikz [remember picture,overlay]
    \node[yshift=-.25cm,xshift=0cm] at (current page.center)
        {\includegraphics[width=.9\paperwidth]{../images/eia_lcoe.png}}; \vspace{1cm}
   %\\[\abovecaptionskip]  \vspace{5.5cm}
   %\tiny{See for more information: \url{https://www.lazard.com/media/451086/lazards-levelized-cost-of-energy-version-130-vf.pdf}}
   \vfill
\end{frame}

\begin{frame}{Costs of New Capacity Additions}
   \tikz [remember picture,overlay]
    \node[yshift=-.25cm,xshift=0cm] at (current page.center)
        {\includegraphics[width=.9\paperwidth]{../images/lazard_lcoe.png}}; \vspace{1cm}
   %\\[\abovecaptionskip]  \vspace{5.5cm}
   %\tiny{See for more information: \url{https://www.lazard.com/media/451086/lazards-levelized-cost-of-energy-version-130-vf.pdf}}
   \vfill
\end{frame}

\begin{frame}{New Technology}
   \tikz [remember picture,overlay]
    \node[yshift=-.5cm,xshift=0cm] at (current page.center)
        {\includegraphics[width=.9\paperwidth]{../images/ab_solar.png}}; \vspace{1cm}
   \vfill
\end{frame}

\begin{frame}{Price Capture by Technology}
   \tikz [remember picture,overlay]
    \node[yshift=-.5cm,xshift=0cm] at (current page.center)
        {\includegraphics[width=.9\paperwidth]{../images/price_capture.png}}; \vspace{1cm}
   \vfill
\end{frame}

\begin{frame}{Evolution of Technology}
   \tikz [remember picture,overlay]
    \node[yshift=-.5cm,xshift=0cm] at (current page.center)
        {\includegraphics[width=.9\paperwidth]{../images/solar_eia.png}}; \vspace{1cm}
   \vfill
\end{frame}

\begin{frame}{Evolution of Technology}
   \tikz [remember picture,overlay]
    \node[yshift=-.5cm,xshift=0cm] at (current page.center)
        {\includegraphics[width=.9\paperwidth]{../images/solar_costs.png}}; \vspace{1cm}
   \vfill
   \vspace{4cm}\tiny{Source: NREL}  \hfill
\end{frame}

\begin{frame}{GHG Policy and Electricity Supply}
   \tikz [remember picture,overlay]
    \node[yshift=-.5cm,xshift=0cm] at (current page.center)
        {\includegraphics[width=.95\paperwidth]{../images/monthly_ghgs.png}}; \vspace{1cm}
   \vfill
\end{frame}

@@

\begin{frame}{Readings and guest speaker}
\begin{itemize}
\setlength\itemsep{2em}
\item Rivers and Dolter paper on decarbonizing Canada's supply
\item van de Biezenbos paper on transmission.

>>>>

\documentclass[aspectratio=169]{beamer}

@@

% Bayliss Fancy fit image command with optional caption
\usepackage{adjustbox} % Shrink stuff

\makeatletter

\newcommand{\fitimage}[2][\@nil]{
  \begin{figure}
    \begin{adjustbox}{width=0.95\textwidth, totalheight=\textheight-2\baselineskip-2\baselineskip,keepaspectratio}
      \includegraphics{#2}
    \end{adjustbox}
    \def\tmp{#1}%
   \ifx\tmp\@nnil
      \else
      \newline{\tiny{#1}}
    \fi
  \end{figure}
}

\makeatother

@@

\begin{frame}{Why aspiring lawyers should care about electricity and what you need to know}
\begin{itemize}
\setlength\itemsep{.5em}
\item Regulatory complexity:
\begin{itemize}
\setlength\itemsep{.5em}
    \item Complex regulation, even in \textit{deregulated} jurisdictions
    \item Each part of the market has its own set of regulatory constructs (distribution, transmission, generation, retail)
\end{itemize}
\item New technology means evolving market structures:
\begin{itemize}
\setlength\itemsep{.5em}
\item Electricity is, arguably, changing faster than any other energy market
\item Alberta's electricity market in the midst of a period of market- and regulatory-driven transition
\end{itemize}
\item Economics 101 in action
\begin{itemize}
\setlength\itemsep{.5em}
\item Supply and demand curves determine prices every hour in Alberta's power market
\end{itemize}
\item The next constitutional battleground?
\begin{itemize}
\setlength\itemsep{.5em}
\item Fed/prov issues, environmental policy, section 125 jurisprudence, the oddity that is section 92(10) including the declaratory power, and POGG all come into play!
\end{itemize}
\end{itemize}
\vfill
\end{frame}

@@

\begin{tabular}{p{\linewidth}}
    \centering
    \includegraphics[width=.75\linewidth]{../images/grid_storage.png} \\[\abovecaptionskip]
  Source: \url{http://www.mpoweruk.com/grid_storage.htm}
\end{tabular}

@@

\begin{frame}{Market Participants}
\begin{itemize}
\setlength\itemsep{1.2em}
\item Generation

@@

\begin{frame}{Market Regulation in Alberta}
\begin{itemize}
\setlength\itemsep{1.7em}
\item Generation is a competitive market

@@

\item Distribution (local wires) is regulated on a cost-of-service basis, but retail (billing) is competitive.

@@

\begin{frame}{Energy units - electricity}
\begin{itemize}
\setlength\itemsep{.75em}
\item Watts: measure of capacity (instantaneous production, installed capacity, or instantaneous demand)
\begin{itemize}
\setlength\itemsep{.5em}
\item Alberta system demand: 7,200-10,700 MW (megawatts or million watts)
\item Capital Power's Genessee 3 power plant has a nameplate capacity of 450 MW
\end{itemize}
\item Watt hours: measure of energy (production or demand during a given period of time; i.e. flow through)
\begin{itemize}
\setlength\itemsep{.5em}
\item Production over a day, week, month, year
\item A 7.5W LED bulb on all day (24hr) uses 180Wh of electricity (.18kWh)
\end{itemize}
\item Volts: measure of the electrical potential or the ability to convert charge to power (Watts=amps x volts)
\begin{itemize}
\setlength\itemsep{.5em}
\item Transmission lines: 150-765 kV
\item Distribution lines: 13,800 Volts
\item Household wiring: 120-240 Volts
\end{itemize}
\end{itemize}
\vfill \end{frame}

@@

\begin{frame}{Regulatory characteristics}
\begin{itemize}
\setlength\itemsep{1.25em}
\item Rate-regulated or or state-owned utilities
\begin{itemize}
\setlength\itemsep{.15em}
\item EPCOR (legacy) or BC Hydro
\item PG \& E in California
\end{itemize}

\item Competitive markets
\begin{itemize}
\setlength\itemsep{.15em}
\item Energy only markets: ERCOT and Alberta
\item Energy and capacity markets: MISO, PJM
\item Real-time vs day-ahead prices: PJM and others have day-ahead market and then a real-time differences market
\item Many other design characteristic differences between restructured or competitive markets
\end{itemize}
\end{itemize}
\vfill
\end{frame}

\begin{frame}{Alberta Wholesale Energy Market Design}
\begin{itemize}
\setlength\itemsep{.25em}
\item Energy-only market
\item Real time, spot pricing, no day-ahead market
\item Single node
\item Capacity market was contemplated, but not pursued
\item Transmission
\begin{itemize}
\setlength\itemsep{.25em}
\item Regulated monopoly
\item Congestion free (no nodal pricing)
\item No transmission rights
\item Paid for (mostly) by load (consumers)
\end{itemize}
\item Ancillary services: separate, competitive market for operating reserves, transmission-must-run, load-shed and black start
\end{itemize}

@@

\begin{tabular}{p{\linewidth}}
    \centering
    \includegraphics[width=.7\linewidth]{../images/cali_nodes.png} \\[.1\abovecaptionskip]
  Source: CAISO
\end{tabular}
\vfill
\end{frame}

@@

\begin{frame}{The Merit Order}
   \fitimage{../images/merit_type.png}
   \vfill
   \end{frame}

\begin{frame}{The Merit Order}
   \fitimage{../images/merit_offer.png}
   \vfill
\end{frame}

\begin{frame}{Offers}
   \fitimage{../images/bids.png}
   \vfill
\end{frame}

\begin{frame}{Hourly Loads}
   \fitimage{../images/hourly_loads.png}
   \vfill
\end{frame}

\begin{frame}{Peak and Average Loads}
   \fitimage{../images/loads_clean.png}
   \vfill
\end{frame}

\begin{frame}{Hourly Prices}
   \fitimage{../images/hourly_prices.png}
   \vfill
\end{frame}

\begin{frame}{Prices over time}
  \fitimage{../images/peak_prices.png}
   \vfill
\end{frame}

\begin{frame}{Forward Markets}
   \fitimage{../images/forwards.png}
   \vfill
\end{frame}

@@

\begin{frame}{The Balancing Pool: What on earth does/did it do?}
\fitimage[Source: Capital Power]{../images/balancingpool.png}
\vfill \end{frame}

@@

\begin{frame}{Costs of New Capacity Additions}
   \fitimage{../images/lcoe_22.png}
   %\\[\abovecaptionskip]  \vspace{5.5cm}
   %\tiny{See for more information: \url{https://www.lazard.com/media/451086/lazards-levelized-cost-of-energy-version-130-vf.pdf}}
   \vfill
\end{frame}

\begin{frame}{Costs of New Capacity Additions}
   \fitimage{../images/lcoe_lazard_22.png}
   \vfill
\end{frame}

\begin{frame}{New Technology}
   \fitimage{../images/ab_solar.png}
   \vfill
\end{frame}

\begin{frame}{New Technology}
   \fitimage{../images/solar.png}
   \vfill
\end{frame}

\begin{frame}{New Technology}
   \fitimage{../images/solar_year.png}
   \vfill
\end{frame}

\begin{frame}{New Technology}
   \fitimage{../images/small_solar.png}
   \vfill
\end{frame}

\begin{frame}{Price Capture by Technology}
\fitimage{../images/price_capture_renew.png}
   \vfill
\end{frame}

\begin{frame}{Evolution of Technology}
   \fitimage{../images/eia_solar.png}
   \vfill
\end{frame}

\begin{frame}{Evolution of Technology}
   \fitimage{../images/eia_coal.png}
   \vfill
\end{frame}

\begin{frame}{Evolution of Technology}
   \fitimage[Source: Lazard]{../images/lazard_decline.png}
   \vfill
\end{frame}

\begin{frame}{GHG Policy and Electricity Supply}
   \fitimage{../images/monthly_ghgs.png}
   \vfill
\end{frame}

@@

\begin{frame}{Readings and guest speaker}
\begin{itemize}
\setlength\itemsep{2em}
\item Rivers and Dolter paper on decarbonizing Canada's supply
\item van de Biezenbos paper on transmission

\item Clean Electricity Standard discussion paper
